Parsing Postfix log files

### 2009-01-14 doc/logparser-other-parsers-review.tex
Further corrections and improvements from proofreading.

XXX DO ANY OF THE PARSERS DISTINGUISH BETWEEN DIFFERENT TYPES OF
REJECTIONS\@?

@@

XXX SHOULD THIS BE IN THE CONCLUSION INSTEAD\@?

@@

    \item A minor difference is that most parsers do not handle compressed
        files; both \parsername{} and Splunk handle them transparently,
        without user intervention; Sawmill and Lire can be configured to
        support compressed files, but Sawmill exhibits a dramatic increase
        in parsing time when doing so.  Although this is a minor
        disparity, support for reading compressed log files is quite
        helpful, as it dramatically reduces the disk space required to
        store historical log files.

    \item Some of the parsers reviewed save the extracted data to a data
        store, but the majority discard all data once they have finished
        generating their report, making historical analysis impossible
        without parsing all log files every time.

@@

Each of the reviewed parsers was tested with the \numberOFlogFILES{} test
log files described in \sectionref{parser efficiency}.  The data extracted
by \parsername{} is documented in \sectionref{connections table} and
\sectionref{results table}; for convenience that list is repeated here:
server \gls{IP}, server hostname, client \gls{IP}, client hostname, HELO
hostname, queueid, start time, end time, \gls{SMTP} code, sender,
recipient, size, message ID\@.

@@

There only appears to be one prior published paper about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \gls{LMA} is quite
different from \parsername{}: it attempts to present correlated data from
log files in a form suitable for a systems administrator to search using
the myriad of standard Unix text processing utilities already available.
It produces a \gls{CSV} file and either a MySQL
(\url{http://www.mysql.com/}) or Berkeley DB
(\url{http://www.oracle.com/database/berkeley-db/index.html}) database.
The decision to support both \gls{CSV} and Berkeley DB appears to have been
a serious limitation: XXX EXTEND\@: LIMITATIONS WILL BE EXPLAINED LATER OR
SOMETHING\@.  Very little documentation is provided with \gls{LMA}, though
some documentation is available in~\cite{log-mail-analyser}.  Studying the
source code is informative, though this author had difficulty as the
authors of \gls{LMA} wrote in Italian.

@@

Pflogsumm produces a report designed for troubleshooting rather than
in-depth analysis.  It does not support saving any data, and it does not
extract any data that it does not require to produce its report, e.g.\ it
does not extract the HELO hostname, queueid, start and end times,
timestamps for each log line, or message id.  Both the parsing and
reporting are difficult to extend because it is a specialised tool, unlike
the easily extensible design of \parsername{}.  It does not correlate log
lines by queueid or \gls{pid}, and does not need to deal with the
complications encountered during this project (\sectionref{complications}).
Pflogsumm produces a useful report, and successfully parsed the
\numberOFlogFILES{} log files it was tested with.\footnote{The results it reported
were not verified in detail, but it did not report any errors, and has a
very good reputation amongst Postfix users.}
Pflogsumm has many options to include or exclude certain
sections of the report; by default it includes the following:

@@

@@

    \item \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline{}
        Fields extracted: from, to, server, UID, relay, status, number of
        recipients, origin hostname, origin \gls{IP}, and virus.  It also
        counts the number of and total size of all mails delivered.  The
        fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
        \texttt{virus} are not explained in the documentation:
        \texttt{server} is probably the hostname or \gls{IP} address of the
        server the mail is delivered to; \texttt{relay} might be the
        delivery method: \gls{SMTP}, local delivery, or \gls{LMTP};
        \texttt{uid} might be the uid of the user submitting mail locally.
        Postfix does not perform any form of virus checking (though it has
        many options for cooperating with an external virus scanner), so
        the \texttt{virus} field is a mystery.

    \item \url{http://www.thesawmill.co.uk/formats/postfix_ii.html}
        \newline{} Fields extracted: from, to, RBL list, client hostname,
        and client \gls{IP}\@.  It also counts the number and total size of
        all mails delivered.

    \item \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
        \newline{} Fields extracted: from, to, client hostname, client
        \gls{IP}, relay hostname, relay \gls{IP}, status, response code,
        RBL list, and message id.  It also counts the number and size of
        all mails delivered, processed, blocked, expired, and bounced.

@@

The source code is available in an obfuscated form, and the product is
quite expensive, requiring a \euros{100} + VAT licence per report
(discounts are available when buying multiple licences); in contrast
\parsername{} is free to use and the code is freely available.  Sawmill is
supplied with thorough and well written documentation; everything the
author searched for was documented, except the MySQL database schema.  A
commercial version of MySQL is required due to MySQL licensing
restrictions, but Sawmill's documentation explains why and includes
instructions on how to compile Sawmill so that it can use a non-commercial
version of MySQL (this was not attempted during the review process).

@@

Isoqlog's report misses most of the information gathered by \parsername{}:
the data extracted is limited to the number of mails sent by each sender,
and it only reports on senders from the domains listed in its configuration
file, making it impossible to produce complete reports.  It ignores all log
lines except those with today's date, so it is impossible to analyse
historical log files, and testing with the \numberOFlogFILES{} test log
files was pointless.  It does maintain a record of data previously
extracted, which the newly extracted data is merged into; the format of the
data store is undocumented.  It does not utilise rejection log lines in any
way, so is unsuitable for the purposes of this project.  Its parsing is
completely inextensible, indeed is almost incomprehensible, relying on
\texttt{scanf(3)}, unexplained fixed offsets, and low level string
manipulation; it is the opposite end of the spectrum to \parsernames{}
parsing.  It does not handle any of the complications discussed in
\sectionref{complications}, does not gather the breadth of data required
for this project, and ignores the majority of log lines produced by
Postfix.

XXX DOCUMENTATION\@?

@@

\noindent{}YAALA's Postfix parser extracts some of the fields \parsername{}
does: for client and server it stores either the \gls{IP} address or the
hostname, not both; it omits the HELO hostname, queueid, \gls{SMTP} code,
size of each accepted mail, start and end times, timestamps for each log
line, and message ID\@.  It extracts some data that \parsername{} does not:
the delay in delivering each mail, and how many times delivery was deferred
for each mail; these could easily be extracted by \parsername{} if desired.
XXX DELAY AND DEFERRAL\@.  Unlike \parsername{}, YAALA does not maintain
separate counters for each restriction; this rules out the possibility of
using the collected data for optimisation, testing, or understanding of
restrictions.  YAALA's Postfix parser does not deal with the complications
explained in \sectionref{complications}, though it does correlate log lines
by queueid.

@@

Lire supports multiple output formats for generated reports (text, HTML,
PDF, and Excel 95) but the reports do not appear to be customisable;
\parsername{} does not produce any reports.  Lire's report contains less
detail than Pflogsumm, and it is considerable harder to configure.  Lire
supports saving extracted data for later report generation, but the format
of this data store is undocumented; given the source code it should be
possible, with enough time and effort, to understand the format.
\parsername{} uses an \gls{SQL} database to make accessing the extracted
data as easy as possible.

@@

It also counts the number of log lines parsed and skipped.  It omits client
\gls{IP} address and hostname, server \gls{IP}, HELO hostname, queueid,
timestamps of each log line, and message ID\@.  It extracts the delay for
delivered mails, which \parsername{} does not.  Log lines are correlated
based on the queueid (referred to as sessionname [sic] within Logrep), but
not by \gls{pid}.  The parsing is error prone: empty fields are saved when
the log line does not match the regex, though it appears that they will not
overwrite existing data.  Most notably rejections are completely ignored,
making it unsuitable for the purposes of this project.  It does not try to
address any of the complications in \sectionref{complications} except for
correlating by queueid.

@@

There are other programs available which perform basic Postfix log file
parsing (some to a greater level of detail than others).   None of the
reviewed parsers collect the breadth of information gathered by
\parsername{}, or make it as easy to extend the parser to handle new log
lines.  Some correlate log lines by queueid (none correlate by pid); none
deal with any of the other complications described in XXX REFERENCE
complications.  All of the reviewed parsers generate a report, and some
provide a greater or lesser degree of customisation.  Most have a data
store, but only \gls{LMA} provides any documentation on its format; some
deliberately make the data store inaccessible to other tools.  Most but not
all of the parsers provide documentation, with the quality varying from
unusable to very good.  Fewer than half of the parsers were capable of
parsing the \numberOFlogFILES{} test log files, and improving or extending
parsing would have been quite a difficult task for any of the parsers.
Table \refwithpage{Summary of parsers' features} provides a summary of the
parsers' features.

XXX REPHRASE The overriding difference between \parsername{} and the other
parsers reviewed herein is that none of them aim for the high level of
understanding of Postfix log files achieved by \parsername{}.

\begin{table}[htb]
    \caption{Summary of parsers' features}
    \empty{}\label{Summary of parsers' features}
    \begin{tabular}{llllll}
        \tabletopline{}%
        Parser          & Parsed test   & Data              & Custom            & Documentation  & Source       \\
                        & log files?    & store?            & reports?          & quality?       & code?        \\
        \tablemiddleline{}%
        \gls{LMA}       & No            & Yes               & No                & Poor           & Yes          \\ 
        Pflogsumm       & Yes           & No                & Partial \dag{}    & Good           & Yes          \\
        Sawmill         & Yes           & Yes               & Searches          & Very good      & \nialpha{}   \\
        Splunk          & Yes           & Yes               & Searches          & Abundant       & No           \\
                        &               &                   & \& reports        & but poor       &              \\
        Isoqlog         & No            & Yes               & No                & No English     & Yes          \\
                        &               &                   &                   & documentation  &              \\
        AWStats         & Partially     & Yes               & Partial \dag{}    & Good           & Yes          \\
        Anteater        & No            & No                & No                & Poor           & Yes          \\
        YAALA           & No            & Yes \ddag{}       & Searches          & Poor           & Yes          \\
        Lire            & Yes           & Yes               & Yes               & Reasonable     & Yes          \\
        Logrep          & No            & Yes               & No                & None           & Yes          \\
        \parsername{}   & Yes           & Yes \nibeta{}     & No \nichi{}       & XXX            & Yes          \\
        \tablebottomline{}%
    \end{tabular}

@@

>>>>

@@

XXX SHOULD THIS LIST BE IN THE CONCLUSION INSTEAD\@?

@@

    \item A minor difference is that most parsers do not handle compressed
        files; both \parsername{} and Splunk handle them transparently,
        without user intervention; Sawmill and Lire can be configured to
        support compressed files, but Sawmill exhibits a dramatic increase
        in parsing time when doing so.  Support for reading compressed log
        files is quite helpful, as it dramatically reduces the disk space
        required to store historical log files.

    \item Most of the reviewed parsers do not distinguish between different
        delivery attempt rejections, so they can not be used to determine
        the success rate of different anti-spam techniques.  The exception
        is Pflogsumm, which provides a summary of why delivery attempts
        were rejected.

@@

Each of the reviewed parsers was tested with the \numberOFlogFILES{} test
log files described in \sectionref{parser efficiency}.  The data extracted
by \parsername{} is documented in \sectionref{connections table} and
\sectionref{results table}; for convenience that list is repeated here:
client and server \gls{IP} address and hostname, HELO hostname, queueid,
start time, end time, \gls{SMTP} code, sender, recipient, size, message
ID\@.

@@

There only appears to be one prior published paper about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \gls{LMA} is quite
different from \parsername{}: it attempts to present correlated data from
log files in a form suitable for a systems administrator to search using
the myriad of standard Unix text processing utilities already available.
It produces a \gls{CSV} file and either a MySQL
(\url{http://www.mysql.com/}) or Berkeley DB
(\url{http://www.oracle.com/database/berkeley-db/index.html}) database.
The decision to support both \gls{CSV} and Berkeley DB appears to have been
a serious limitation: both formats have limitations that will be explained
later.  Very little documentation is provided with \gls{LMA}, though some
documentation is available in~\cite{log-mail-analyser}.  Studying the
source code is informative, though this author had difficulty as the
authors of \gls{LMA} wrote in Italian.

@@

Pflogsumm produces a report designed for troubleshooting rather than
in-depth analysis.  It does not support saving any data, and it does not
extract any data that it does not require to produce its report, e.g.\ it
does not extract the HELO hostname, queueid, start and end times,
timestamps for each log line, or message id.  Both the parsing and
reporting are difficult to extend because it is a specialised tool, unlike
the easily extensible design of \parsername{}.  It does not correlate log
lines by queueid or \gls{pid}, and does not need to deal with the
complications encountered during this project (\sectionref{complications}).
Pflogsumm produces a useful report, and successfully parsed the
\numberOFlogFILES{} log files it was tested with.\footnote{The results it
reported were not verified in detail, but it did not report any errors, and
has a very good reputation amongst Postfix users.}  Pflogsumm has many
options to include or exclude certain sections of the report, all clearly
documented; by default it includes the following:

@@

Pflogsumm is the only reviewed parser which distinguishes between different
rejections.

@@

    \item \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline{}
        Fields extracted: from, to, server, UID, relay, status, number of
        recipients, origin hostname, origin \gls{IP} address, and virus.
        It also counts the number of and total size of all mails delivered.
        The fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
        \texttt{virus} are not explained in the documentation:
        \texttt{server} is probably the hostname or \gls{IP} address of the
        server the mail is delivered to; \texttt{relay} might be the
        delivery method: \gls{SMTP}, local delivery, or \gls{LMTP};
        \texttt{uid} might be the uid of the user submitting mail locally.
        Postfix does not perform any form of virus checking (though it has
        many options for cooperating with an external virus scanner), so
        the \texttt{virus} field is a mystery.

    \item \url{http://www.thesawmill.co.uk/formats/postfix_ii.html}
        \newline{} Fields extracted: from, to, RBL list, client hostname,
        and client \gls{IP} address.  It also counts the number and total
        size of all mails delivered.

    \item \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
        \newline{} Fields extracted: from, to, client hostname, client
        \gls{IP} address, relay hostname, relay \gls{IP} address, status,
        response code, RBL list, and message id.  It also counts the number
        and size of all mails delivered, processed, blocked, expired, and
        bounced.

@@

The source code is available in an obfuscated form, and the product is
quite expensive, requiring a \euros{100} + VAT licence per report
(discounts are available when buying multiple licences); in contrast
\parsername{} is free to use and the code is freely available.  Sawmill is
supplied with thorough and well written documentation; everything the
author searched for was documented, except the MySQL database schema and
some details of the data extracted by the parser.  A commercial version of
MySQL is required due to MySQL licensing restrictions, but Sawmill's
documentation explains why and includes instructions on how to compile
Sawmill so that it can use a non-commercial version of MySQL (this was not
attempted during the review process).

@@

Isoqlog's report misses most of the information gathered by \parsername{}:
the data extracted is limited to the number of mails sent by each sender,
and it only reports on senders from the domains listed in its configuration
file, making it impossible to produce complete reports.  It ignores all log
lines except those with today's date, so it is impossible to analyse
historical log files, and testing with the \numberOFlogFILES{} test log
files was pointless.  It does maintain a record of data previously
extracted, which the newly extracted data is merged into; the format of the
data store is undocumented.  Very little documentation is provided with
Isoqlog, little more than installation instructions.  It does not utilise
rejection log lines in any way, so is unsuitable for the purposes of this
project.  Its parsing is completely inextensible, indeed is almost
incomprehensible, relying on \texttt{scanf(3)}, unexplained fixed offsets,
and low level string manipulation; it is the opposite end of the spectrum
to \parsernames{} parsing.  It does not handle any of the complications
discussed in \sectionref{complications}, does not gather the breadth of
data required for this project, and ignores the majority of log lines
produced by Postfix.

@@

\noindent{}YAALA's Postfix parser extracts some of the fields \parsername{}
does: for client and server it stores either the \gls{IP} address or the
hostname, not both; it omits the HELO hostname, queueid, \gls{SMTP} code,
size of each accepted mail, start and end times, timestamps for each log
line, and message ID\@.  It extracts some data that \parsername{} does not:
the delay in delivering each mail, and how many times delivery was deferred
for each mail; these could easily be extracted by \parsername{} if desired.
Unlike \parsername{}, YAALA does not maintain separate counters for each
restriction; this rules out the possibility of using the collected data for
optimisation, testing, or understanding of restrictions.  YAALA's Postfix
parser does not deal with the complications explained in
\sectionref{complications}, though it does correlate log lines by queueid.

@@

Lire supports multiple output formats for generated reports (text, HTML,
PDF, and Excel 95) but the reports do not appear to be customisable;
\parsername{} does not produce any reports.  Lire's report contains less
detail than Pflogsumm, and it is considerable harder to configure.  Lire
supports saving extracted data for later report generation, but the format
of this data store is undocumented; given the source code it should be
possible, with enough time and effort, to understand the format.
\parsername{} uses an \gls{SQL} database to make accessing the extracted
data as easy as possible.  In general Lire has poor documentation.

@@

It also counts the number of log lines parsed and skipped.  It omits client
\gls{IP} address and hostname, server \gls{IP} address, HELO hostname,
queueid, timestamps of each log line, and message ID\@.  It extracts the
delay for delivered mails, which \parsername{} does not.  Log lines are
correlated based on the queueid (referred to as sessionname [sic] within
Logrep), but not by \gls{pid}.  The parsing is error prone: empty fields
are saved when the log line does not match the regex, though it appears
that they will not overwrite existing data.  Most notably rejections are
completely ignored, making it unsuitable for the purposes of this project.
It does not try to address any of the complications in
\sectionref{complications} except for correlating by queueid.

@@

There are other programs available which perform basic Postfix log file
parsing (some to a greater level of detail than others).   None of the
reviewed parsers collect the breadth of information gathered by
\parsername{}, or make it as easy to extend the parser to handle new log
lines.  Some correlate log lines by queueid (none correlate by pid); none
deal with any of the other complications described in
\sectionref{complications}.  All of the reviewed parsers generate a report,
and some provide a greater or lesser degree of customisation.  Most have a
data store, but only \gls{LMA} provides any documentation on its format;
some deliberately make the data store inaccessible to other tools.  Most
but not all of the parsers provide documentation, with the quality varying
from unusable to very good.  Fewer than half of the parsers were capable of
parsing the \numberOFlogFILES{} test log files, and improving or extending
parsing would have been quite a difficult task for any of the parsers.
Table \refwithpage{Summary of parsers' features} provides a summary of the
parsers' features.  The overriding difference between \parsername{} and the
other parsers reviewed herein is that none of them aim for the high level
of understanding of Postfix log files achieved by \parsername{}.

\begin{table}[htb]
    \caption{Summary of parsers' features}
    \empty{}\label{Summary of parsers' features}
    \begin{tabular}{llllll}
        \tabletopline{}%
        Parser          & Parsed test   & Data              & Custom            & Documentation  & Source       \\
                        & log files?    & store?            & reports?          & quality?       & code?        \\
        \tablemiddleline{}%
        \gls{LMA}       & No            & Yes               & No                & Poor           & Yes          \\ 
        Pflogsumm       & Yes           & No                & Partial \dag{}    & Good           & Yes          \\
        Sawmill         & Yes           & Yes               & Searches          & Very good      & \nialpha{}   \\
        Splunk          & Yes           & Yes               & Searches          & Abundant       & No           \\
                        &               &                   & \& reports        & but poor       &              \\
        Isoqlog         & No            & Yes               & No                & No English     & Yes          \\
                        &               &                   &                   & documentation  &              \\
        AWStats         & Partially     & Yes               & Partial \dag{}    & Good           & Yes          \\
        Anteater        & No            & No                & No                & None           & Yes          \\
        YAALA           & No            & Yes \ddag{}       & Searches          & Poor           & Yes          \\
                        &               &                   & \& reports        &                &              \\
        Lire            & Yes           & Yes               & Yes               & Reasonable     & Yes          \\
        Logrep          & No            & Yes               & No                & None           & Yes          \\
        \parsername{}   & Yes           & Yes \nibeta{}     & No \nichi{}       & \niepsilon{}   & Yes          \\
        \tablebottomline{}%
    \end{tabular}

@@

        \item [\niepsilon{}] \parsername{} aims to have thorough and
            complete documentation, but the author cannot provide an
            unbiased review.
